---
author: "Bernhard Scheliga"
date: "2020-09-04"
output: pdf_document
title: "R Notebook for SAPE 2019"
---


Data set created by  
Author: Bernhard Scheliga 
```{r echo=FALSE}
#Enter the Version of the data set here
Dataset_version <- "0.1"
```

```{r echo=FALSE}
print(paste("Dataset version:", Dataset_version, sep = " "))
```

```{r echo=FALSE}
print(paste("Date:", Sys.Date()))
```
```{r echo=FALSE}
version$version.string

```

### 1. Summary:
Initially, the plan was to included this data directly in the Scottish Vulnerability Resource (SVR). However, the Small Area Population Estimate (SAPE) 2019 data set for Scotland is a bit too large in itself to be included directly. The idea now is, that it is an additional resource to the SVR and we will provide a script for user to combine the SVR and the SVR_SAPE2019 dataset. 

The SVR_SAPE2019 resource is postcode(PC) searchable and enables the user to swiftly retrieve data for datazones based on a postcode search. *!Data is not broken down to postcode level!* The spatial resolution of the resource are datazone. For more details see our in depth ***LINK TO THE RIGHT FILE ***[documentation](https://github.com/AbdnCHDS/Scotland_Vulnerability_Resource/blob/master/Documentation/Scotland%20vulernability%20data%20documentation.pdf) on [github.com/AbdnCHDS/Scotland_Vulnerability_Resource](https://github.com/AbdnCHDS/Scotland_Vulnerability_Resource)

Here we describe how the SVR_SAPE2019 was created in *R*, which code and information was used. This is to enable the interested user to follow our process and get a better understanding of the decisions we during the creation process of the SVR_SAPE2019. 

The SVR_SAPE2019 is build around the Small Area Population Estimate (SAPE) 2019 data set for Scotland. The current version of the SVR_SAPE 2019 include Female, Male and Both population estimates, the postcodes (PC) of the respective SIMD2020v2 data zones, their data zone names adn the NHS Health board regions. 


Keywords: *R script, SAPE 2019, SIMD2020v2, postcode searchable, datazone names, reproducible, open access*

### 2. Creating the data set  
#### 2.1 Loading source data  


```{r echo=TRUE}
setwd("~/Scotland_Vulnerability_Resource/SVR-data/")
dir()
df_SVR <- read.csv("Scotland-Vulnerability-Resource_v0.2.csv")# we only want the first few columns 


## The SAPE 2019 Data
setwd("~/Scotland_Vulnerability_Resource/Input-data/")
dir()
df_SAPE2019.Female <-read.csv("sape-2019-females_Table 1c Females (2019).csv", skip = 3)
df_SAPE2019.Male <- read.csv("sape-2019-males_Table 1b Males (2019).csv", skip = 3)
#df_SAPE2019.Person <-read.csv("sape-2019-persons_Table 1a Persons (2019).csv", skip = 3)

library("tidyverse")
```
#### 2.2 Cleaning source data 
##### 2.2.1 Removing excess data from source data 

```{r}
df_SAPE2019.Female %>% head()
```
This is what all SAPE 2019 data set look like. THe first two row and last two rows we do not need. Row 1 has the total poptulation in Scotland for the respective gender and age group. Row 2 has only the header for for the first 3 columns.
We will remove column X.1, X.2, X.3, X.4, X.5 & X.6


```{r}
df_SAPE2019.Female <- df_SAPE2019.Female %>% select(-c(X.1, X.2, X.3, X.4, X.5, X.6)) %>% slice(-c(1,2,6979,6980))
df_SAPE2019.Male <- df_SAPE2019.Male %>% select(-c(X.1, X.2, X.3, X.4, X.5, X.6)) %>% slice(-c(1,2,6979,6980))
#df_SAPE2019.Person <- df_SAPE2019.Person %>% select(-c(X.1, X.2, X.3, X.4, X.5, X.6)) %>% slice(-c(1,2,6979,6980))

```

We need to change two column names, see below

```{r}
colnames(df_SAPE2019.Female)[c(1,92)]
```

```{r}

colnames(df_SAPE2019.Female)[c(1,92)] <- c("Data_Zone","AGE90PLUS")
colnames(df_SAPE2019.Male)[c(1,92)] <- c("Data_Zone","AGE90PLUS")
#colnames(df_SAPE2019.Person)[c(1,92)] <- c("Data_Zone","AGE90PLUS")

colnames(df_SAPE2019.Female)[c(1,92)]
```
Better.

##### Adding a gender column

This is done to distinglish the data later once it is merged

```{r}
df_SAPE2019.Female$Gender <- "Female"
df_SAPE2019.Male$Gender <- "Male"
#df_SAPE2019.Person$Gender <- "Person"
```

#### 3. Joining the Datasets

We actually just need to use bind_rows() the datasets.
***Note!***: We only joined the female & male data since the all three datasets together would be to big for GitHub.

```{r}
df_SAPE2019 <- bind_rows(df_SAPE2019.Female,df_SAPE2019.Male)

```

Now, we have all three SAPE 2019 data set in one object. The next step is to take the *Postcode -, Data_zone -, Intermediate_Zone -,Council_area - & NHS_Health_Board_Region -* columns from the SVR dataset and join them to the SAPE 2019 (*df_SAPE2019*).

```{r}
df_SAPE2019.SVRfront <- df_SVR %>% select(Postcode, Data_Zone, Intermediate_Zone,Council_area, NHS_Health_Board_Region) %>% inner_join(df_SAPE2019, by= "Data_Zone")
# quick reordering of the columns
df_SAPE2019.SVRfront <- df_SAPE2019.SVRfront[,c(1:5,97,6:96)]

## Check for NA
sapply(df_SAPE2019.SVRfront, function(x) sum(is.na(x)))

```
6 NA in Postcode. That will be the same NA postcodes as in the SVR data (Petershill & Sighthill) just time three

```{r}
df_SAPE2019.SVRfront[is.na(df_SAPE2019.SVRfront$Postcode),]
```
We want to check now, if there are only *numeric* value in the *AGE*-columns. By changing the value type from *character* to *numeric* in those columns non-numeric values should turn into NA-values.

```{r}
df_SAPE2019.SVRfront[,c(7:97)] <- sapply(df_SAPE2019.SVRfront[,c(7:97)], as.numeric) # Couldn't find a better way to do this

df_SAPE2019.SVRfront%>%select(AGE0:AGE90PLUS)%>%sapply(function(x) sum(is.na(x)))

```
No NA-values, perfect!


#### 2.4 Saving the data set

```{r}
setwd("~/Scotland_Vulnerability_Resource/SVR-data/")

write.csv(df_SAPE2019.SVRfront, paste("2019_Small_Area_Population_Estimates_FeMale_SVR",".csv", sep = ""), row.names=FALSE)
```
