---
title: "R Notebook for creation Grampian Data set"
output: word_document
---

Rscript created by   
Author: Bernhard Scheliga  
Date: 2020-06-11

Data set created by  
Author: Bernhard Scheliga 
```{r echo=FALSE}
#Enter the Version of the data set here
Dataset_version <- "0.2"
```

```{r echo=FALSE}
print(paste("Dataset version:", Dataset_version, sep = " "))
```

```{r echo=FALSE}
print(paste("Date:", Sys.Date()))
```
```{r echo=FALSE}
version$version.string

```
### 1. Summary:
**The data set name (Grampian data) is misleading at the moment as it has all of Scotland's SIMD data. Sorry**

This script adds additional information to the Scottish Index of Multiple Deprivation indicators (SIMD2020v2) data set. __*Consider adding more details e.g. best on openly accessible information bla and when information access. However, not 100% here would be the right place*__ The current version of the script adds the postcodes (PC) of the respective SIMD2020v2 data zones, their data zone names adn the NHS Health board regions. 

### 2. Creating the data set  
#### 2.1 Loading source data  
```{r echo=TRUE}

setwd("~/grampian_data/Raw_data/")
dir()
df_SIMD2020.indi <- read.csv("SIMD2020v2indicators.csv")
df_SIMD2020.dz <- read.csv("SIMD2020v2datazones.csv")
df_SIMD2020.pc <- read.csv("SIMD2020v2postcodes.csv")
df_NHS_regions <- read.csv("NHS_Health_Board_regions.csv")
```

#### 2.2 Cleaning source data 
##### 2.2.1 Removing excess data from source data  

```{r echo=TRUE}

# df_SIMD.dz we only need column 1,3:13,16,17. See Issues #19 here for details https://github.com/AbdnCHDS/grampian_data/issues/19
df_SIMD2020.dz <- df_SIMD2020.dz[,c(1,3:13,16,17)]
# From df_SIMD2020.pc we only need the first two columns "Postcode" & "DZ"
df_SIMD2020.pc <- df_SIMD2020.pc[,c(1,2)]
```

##### 2.2.2 Checking source data for duplicates  

```{r echo=TRUE}
sapply(df_SIMD2020.indi, function(x) sum(duplicated(x)))
# "Data_Zone" should be 0, as they are our primary key here

sapply(df_SIMD2020.dz, function(x) sum(duplicated(x)))
# "DZ" (Data zone) should be 0, as they are our primary key here

sapply(df_SIMD2020.pc, function(x) sum(duplicated(x)))
# "Postcode" should be 0, as they are our primary key here


```

##### 2.2.3 Checking source data for NA-values  

```{r echo=TRUE}
sapply(df_SIMD2020.indi, function(x) sum(is.na(x)))
# Data_Zone should be 0. However, currently in the SIM2020v2 source data set missing values and suppressed values are denoted by "*" 

# Lets check how many missing values and suppressed values are denoted by "*" in the source data 
sapply(df_SIMD2020.indi, function(x) sum(x=="*"))

sapply(df_SIMD2020.dz, function(x) sum(is.na(x)))
# DZ (Data zone) should be 0
sapply(df_SIMD2020.pc, function(x) sum(is.na(x)))
# Postcode should be 0

```
Nothing concerning here :)


### 2.3 Joining the source data set  

#### 2.3.1 Merging/Joining data zone names to SIMD2020v2
```{r echo=TRUE}
df_SIMD2020.1merge <- merge(df_SIMD2020.indi, df_SIMD2020.dz, by.x="Data_Zone", by.y="DZ", all = TRUE) #all= TRUE to include potential missing values. In case something goes wrong with merge().

# Check if we introduced NA values
sapply(df_SIMD2020.1merge, function(x) sum(is.na(x)))

```

#### 2.3.2 Merging/Joining postcodes to SIMD2020v2

```{r echo=TRUE}

df_SIMD2020.2merge <- merge(df_SIMD2020.1merge, df_SIMD2020.pc, by.x="Data_Zone", by.y="DZ", all = TRUE)
# reordering the data frame, placing the postcode column in the first position
df_SIMD2020.2merge <- df_SIMD2020.2merge[,c(51,1:50)]


# find the column with the NA values
sapply(df_SIMD2020.2merge, function(x) sum(is.na(x)))
# find affected rows
df_SIMD2020.2merge[is.na(df_SIMD2020.2merge$Postcode),]


```
There are no postcodes for those to data zone. Also they don't have any population. See issue #9 on [link](https://github.com/AbdnCHDS/grampian_data/issues).

```{r echo=TRUE}
## duplicated values
sapply(df_SIMD2020.2merge, function(x) sum(duplicated(x)))
# However, there is one duplicated postcode. The Postcodes should be 0
# let find it

df_SIMD2020.2merge[duplicated(df_SIMD2020.2merge$Postcode, fromLast=FALSE),]
# duplicated also picked up on the two NA in the postcodes

```


#### 2.3.2 Merging/Joining NHS Health Board regions to SIMD2020v2  

```{r echo=TRUE}
df_SIMD2020.3merge <- merge(df_SIMD2020.2merge, df_NHS_regions, by.x="Council_area", by.y="Council_area", all = TRUE)
# reordering the data frame
df_SIMD2020.3merge <- df_SIMD2020.3merge[,c(2:4,1,52,5:51)]
# find the column with the NA values
sapply(df_SIMD2020.3merge, function(x) sum(is.na(x)))
# find affected rows
df_SIMD2020.3merge[is.na(df_SIMD2020.3merge$Postcode),]
```


#### 2.4 Saving the data set

```{r}
setwd("~/grampian_data/Processed_data/")

write.csv(df_SIMD2020.3merge, paste("Grampian_data_v",Dataset_version,".csv", sep = ""), row.names=FALSE)
```

