---
title: "R Notebook for creation Scotland Vulnerability Resource"
output: pdf_document
---

Rscript created by   
Author: Bernhard Scheliga  
Date: 2020-06-11

Data set created by  
Author: Bernhard Scheliga 
```{r echo=FALSE}
#Enter the Version of the data set here
Dataset_version <- "0.2"
```

```{r echo=FALSE}
print(paste("Dataset version:", Dataset_version, sep = " "))
```

```{r echo=FALSE}
print(paste("Date:", Sys.Date()))
```
```{r echo=FALSE}
version$version.string

```

### 1. Summary:

The Scottish Vulnerability Resource (SVR) is a composition of different openly accessible datasets. It a ready-to-use open access resource that includes demographic, socioeconomic, and deprivation measures for Scotland. Which allows the user to easily create there own definition of vulnerability and compare those measures within the different regions of Scotland. Additionally, this resource is postcode(PC) searchable and enables the user to swiftly retrieve data for datazones based on a postcode search. *!Data is not broken down to postcode level!* The spatial resolution of the resource are datazone. For more details see our in depth [documentation](https://github.com/AbdnCHDS/Scotland_Vulnerability_Resource/blob/master/Documentation/Scotland%20vulernability%20data%20documentation.pdf) on [github.com/AbdnCHDS/Scotland_Vulnerability_Resource](https://github.com/AbdnCHDS/Scotland_Vulnerability_Resource)

Here we describe how the SVR was created in *R*, which code and information was used. This is to enable the interested user to follow our process and get a better understanding of the decisions we during the creation process of the SVR. 

The SVR is build around the Scottish Index of Multiple Deprivation indicators (SIMD2020v2) as the core dataset. The current version of the SVR adds the postcodes (PC) of the respective SIMD2020v2 data zones, their data zone names adn the NHS Health board regions. 


Keywords: *script, SIMD2020v2, postcode searchable, datazone names, reproducible, open access*

### 2. Creating the data set  
#### 2.1 Loading source data  
```{r echo=TRUE}

setwd("~/Scotland_Vulnerability_Resource/Raw_data/")
dir()
df_SIMD2020.indi <- read.csv("SIMD2020v2indicators.csv")
df_SIMD2020.dz <- read.csv("SIMD2020v2datazones.csv")
df_SIMD2020.pc <- read.csv("SIMD2020v2postcodes.csv")
df_NHS_regions <- read.csv("NHS_Health_Board_regions.csv")
```

#### 2.2 Cleaning source data 
##### 2.2.1 Removing excess data from source data  

```{r echo=TRUE}

# df_SIMD.dz we only need column 1,3:13,16,17. See Issues #19 here for details https://github.com/AbdnCHDS/grampian_data/issues/19
df_SIMD2020.dz <- df_SIMD2020.dz[,c(1,3:13,16,17)]
# From df_SIMD2020.pc we only need the first two columns "Postcode" & "DZ"
df_SIMD2020.pc <- df_SIMD2020.pc[,c(1,2)]
```

##### 2.2.2 Checking source data for duplicates  

```{r echo=TRUE}
sapply(df_SIMD2020.indi, function(x) sum(duplicated(x)))
# "Data_Zone" should be 0, as they are our primary key here

sapply(df_SIMD2020.dz, function(x) sum(duplicated(x)))
# "DZ" (Data zone) should be 0, as they are our primary key here

sapply(df_SIMD2020.pc, function(x) sum(duplicated(x)))
# "Postcode" should be 0, as they are our primary key here


```

##### 2.2.3 Checking source data for NA-values  

```{r echo=TRUE}
sapply(df_SIMD2020.indi, function(x) sum(is.na(x)))
# Data_Zone should be 0. However, currently in the SIM2020v2 source data set missing values and suppressed values are denoted by "*" 

# Lets check how many missing values and suppressed values are denoted by "*" in the source data 
sapply(df_SIMD2020.indi, function(x) sum(x=="*"))

# storing for a quick comparision 
df_Star <-sapply(df_SIMD2020.indi, function(x) sum(x=="*"))

# Here we replace the "*" denotation with NA in the SIMD data set
df_SIMD2020.indi[df_SIMD2020.indi=="*"] <- NA
# Comparing, if number of "*" denotes is the same as the number of NA now for the   respective columns
df_Star == sapply(df_SIMD2020.indi, function(x) sum(is.na(x)))


sapply(df_SIMD2020.dz, function(x) sum(is.na(x)))
# DZ (Data zone) should be 0
sapply(df_SIMD2020.pc, function(x) sum(is.na(x)))
# Postcode should be 0

```
Nothing concerning here :)


### 2.3 Joining the source data set  

#### 2.3.1 Merging/Joining data zone names to SIMD2020v2
```{r echo=TRUE}
df_SIMD2020.1merge <- merge(df_SIMD2020.indi, df_SIMD2020.dz, by.x="Data_Zone", by.y="DZ", all = TRUE) #all= TRUE to include potential missing values. In case something goes wrong with merge().

# Check if we introduced NA values
sapply(df_SIMD2020.1merge, function(x) sum(is.na(x)))

```

#### 2.3.2 Merging/Joining postcodes to SIMD2020v2

```{r echo=TRUE}

df_SIMD2020.2merge <- merge(df_SIMD2020.1merge, df_SIMD2020.pc, by.x="Data_Zone", by.y="DZ", all = TRUE)
# reordering the data frame, placing the postcode column in the first position
df_SIMD2020.2merge <- df_SIMD2020.2merge[,c(51,1:50)]


# find the column with the NA values
sapply(df_SIMD2020.2merge, function(x) sum(is.na(x)))
# find affected rows
df_SIMD2020.2merge[is.na(df_SIMD2020.2merge$Postcode),]


```
There are no postcodes for those to data zone. Also they don't have any population. See issue #9 on [link](https://github.com/AbdnCHDS/grampian_data/issues).

```{r echo=TRUE}
## duplicated values
sapply(df_SIMD2020.2merge, function(x) sum(duplicated(x)))
# However, there is one duplicated postcode. The Postcodes should be 0
# let find it

df_SIMD2020.2merge[duplicated(df_SIMD2020.2merge$Postcode, fromLast=FALSE),]
# duplicated also picked up on the two NA in the postcodes

```


#### 2.3.2 Merging/Joining NHS Health Board regions to SIMD2020v2  

```{r echo=TRUE}
df_SIMD2020.3merge <- merge(df_SIMD2020.2merge, df_NHS_regions, by.x="Council_area", by.y="Council_area", all = TRUE)
# reordering the data frame
df_SIMD2020.3merge <- df_SIMD2020.3merge[,c(2:4,1,52,5:51)]
# find the column with the NA values
sapply(df_SIMD2020.3merge, function(x) sum(is.na(x)))
# find affected rows
df_SIMD2020.3merge[is.na(df_SIMD2020.3merge$Postcode),]
```


#### 2.4 Saving the data set

```{r}
setwd("~/Scotland_Vulnerability_Resource/Processed_data/")

write.csv(df_SIMD2020.3merge, paste("Scotland_Vulnerability_Resource_v",Dataset_version,".csv", sep = ""), row.names=FALSE)
```

